---
title: "Summary"
subtitle: "Summary of Findings From UNICEF MICS Data"
author: "Andrew Grogan-Kaylor"
date: "today"
format: html
editor: source
---

```{r}
#| output: false
#| echo: false

library(ggplot2)

library(readr)

library(plotly)

library(dplyr)

library(countrycode)

```

# Research Question

We began by setting out to examine the associations of *physical punishment* and *verbal reasoning* with *child development outcomes*.

A second research question was: Are the relationships between different types of parenting *consistent* around the world?

```{mermaid}

flowchart LR

  %% Class Definitions %%

  classDef yellow fill:#FFC20E,stroke:#000000,stroke-width:2px,color:#000000;
  
  classDef blue fill:#374EA2,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef green fill:#00833D,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef orange fill:#D86018,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef red fill:#9A3324,stroke:#000000,stroke-width:2px,color:#FFFFFF;

 classDef grey fill:#D8D1C9,stroke:#000000,stroke-width:2px,color:#000000;
    
  %% Nodes %%
 
  explained[explained why behavior was wrong]:::grey 
  
  CP[physical punishment]:::grey
  
  outcome[child development outcomes]:::grey
  
  %% Links %%

  explained --> |"???"| outcome
  
  CP --> |"???"| outcome
  
  %% Links %%
  
  linkStyle 0,1 stroke:#E2231A,stroke-width:4px,color:red, font-size:18pt;

```

# Data 

To carry out our various research projects, we made use of the *Multiple Indicator Cluster Surveys* (MICS) from UNICEF. 

The MICS surveys contain questions on parenting, and child behaviors, and have been carried out in over 60 *Low and Middle Income Countries* (LMICs) around the world. 

```{r}
#| echo: false
#| warning: false

country <- c("Algeria",  "Argentina",  "Bangladesh",  "Barbados",  "Belarus",  "Belize",  "Benin",  "Bosnia and Herzegovina",  "Cameroon",  "Central African Republic",  "Chad",  "Democratic Republic of the Congo",  "Republic of the Congo",  "Costa Rica",  "Cote d'Ivoire",  "Dominican Republic",  "El Salvador",  "Eswatini",  "Ghana",  "Guinea",  "Guinea Bissau",  "Guyana",  "Iraq",  "Jamaica",  "Kazakhstan",  "Kenya",  "Kosovo",  "Kyrgyzstan",  "Laos", "Macedonia",  "Madagascar",  "Malawi",  "Mali",  "Mauritania",  "Mexico",  "Moldova",  "Mongolia",  "Montenegro",  "Nepal",  "Nigeria",  "Pakistan",  "Panama",  "Paraguay",  "Sao Tome and Principe",  "Senegal",  "Serbia",  "Sierra Leone",  "Somalia",  "St. Lucia",  "State of Palestine",  "Suriname",  "Thailand",  "The Gambia",  "Togo",  "Trinidad and Tobago",  "Tunisia",  "Turkmenistan",  "Ukraine",  "Uruguay",  "Vietnam",  "Zimbabwe")

MICScountries <- data.frame(country)

MICScountries$country_iso <- countrycode(MICScountries$country, 
                                 'country.name', 
                                 'iso3c')

```

```{r}
#| echo: false

g <- list(showframe = TRUE, 
          showcoastlines = TRUE, 
          showcountries = TRUE,
          showland = TRUE,
          countrycolor = toRGB("grey"),
          showocean = TRUE,
          oceancolor = "white",
          projection = list(type = 'robinson'),
          showland = TRUE,
          landcolor = toRGB("lightgrey"),
          bgcolor = "transparent")

t <- list(family = "sans serif", 
          size = 8, 
          color = toRGB("black"))

l <- list(color = toRGB("black"), 
          width = 0.5)

p0 <- MICScountries %>% 
  plot_geo() %>% 
  add_trace(z = 1,
    text = ~country, 
    colors = "#1CABE2",
    hoverinfo = 'text',
    locations = ~country_iso, 
    showscale = FALSE,
    marker = list(size=10, line = l)) %>%
  layout(geo = g,
         plot_bgcolor='transparent',
         paper_bgcolor='transparent') %>% 
  config(displayModeBar = FALSE)

```

```{r}
#| echo: false
#| fig-cap: "Countries in MICS"

p0

```

# Major Findings

There are differences in parenting *behaviors* across countries. Encouragingly, the most commonly employed disciplinary technique is explaining (verbal reasoning), but psychologically and physically aggressive forms of discipline are also very common.

These differences in parenting *behaviors* notwithstanding, physical punishment (spanking) is consistently associated with *decreases in child socio-emotional development*, and particularly with *increases in child aggression*.

* There is some variation in the effects of parental discipline (including spanking), on child well-being, but across multiple diverse countries, the *effects* of parental discipline (including spanking) are *remarkably consistent and nearly universal*.
* Country level norms seem to have an association with child outcomes. In countries where spanking is more normative, child aggression is higher. 

Psychological aggression (e.g. calling a child names) may have effects that are as harmful as those of physical punishment.

* Positive discipline--in the form of verbal reasoning--is generally associated with improvements in child well-being across countries, and should therefore be recommended. 
* The association of other forms of nonviolent discipline with child outcomes is less clear. It may be that the long term investments that parents make in children, such as spending time with children, letting children know that they love them, and listening to children have more positive effects than nonviolent discipline, but this has yet to be thoroughly researched in global context.

Simulations with MICS data suggest that eliminating the use of physical punishment would be met with a large reduction in the amount of physical child abuse.

