---
title: "Why Do We Need Multivariate Models To Evaluate Programs?"
author: 
  - name: "Andy Grogan-Kaylor"
    url: https://agrogan1.github.io/
date: "today"
format: 
  html:
    toc: true
    number-sections: true
mermaid:
  theme: neutral
filters:
  - highlight-text
---

# Introduction

Across the world, there is a great deal of suffering. Many people have mental health issues or substance use issues. People often suffer the after effects of discrimination, poverty, trauma, violence or conflict.

```{r}
#| echo: false
#| message: false

library(rnaturalearth) # natural earth data

library(ggplot2) # beautiful maps

library(dplyr) # data wrangling

library(sf) # simple (spatial) features

# library(patchwork) # stitch ggplots together

mapdata <- ne_countries(scale = "medium", # medium scale
                        returnclass = "sf") # as sf object

# CF https://stackoverflow.com/questions/75860459/change-ocean-color-in-robinson-projection-map

ocean <- st_polygon(list(cbind(c(seq(-180, 179, 
                                     len = 100), 
                                 rep(180, 100), 
                                 seq(179, -180, 
                                     len = 100),
                                 rep(-180, 100)),
                               c(rep(-90, 100), 
                                 seq(-89, 89, 
                                     len = 100),
                                 rep(90, 100), 
                                 seq(89, -90, 
                                     len = 100))))) |>
  st_sfc(crs = "WGS84") |>
  st_as_sf()

mymap <- ggplot(mapdata) + # the data I am mapping
  geom_sf(data = ocean, fill = "#ECECFF") +
  geom_sf(fill = "#FFF4DD") + # the geometry I am using
  theme_minimal() +
  theme(axis.text.x = element_blank())

mymap + coord_sf(crs="+proj=robin") # Robinson projection

```

Understandably, many people and organizations try to develop *interventions* or *programs* for those who suffer from such difficulties. 

Yet evaluating such *programs* may be more difficult than it appears.

# A Simple Evaluation

Let's consider a simple evaluation of a *program* designed to improve mental health.

In its simplest form, an evaluation might consist of looking at the outcomes--e.g. mental health outcomes--for those who participate in an *program*.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

  program[program] --> outcome1[outcome]

  linkStyle 0 stroke:#000000,stroke-width:3px,font-size:36px,color:black;

```

# Our Worry

However, we might wonder, or worry, about a number of issues. For example:

* What were the outcomes like for this group of people *before* they participated in the *program*?
* If the outcomes of the *program* were favorable, it might not be that the program is particularly better, but that people improve or get better *naturally* over time. 

# A More Sophisticated Evaluation

Therefore a more sophisticated research design would be to have one group of people--an *program group*--participate in the *program*, while another group--a *comparison* group--does not participate. We would then compare outcomes across the two groups.[^ethics]

[^ethics]: Valid questions could be raised about the *ethics* of such an approach, specifically denying participation in the *program* to one group of people. If a program is of *unknown* benefit, it is *ethical* to evaluate this program with a comparison group approach, because it is *not yet known* whether the program confers benefit on its participants, and whether the program represents a valid use of time, energy and financial resources, or whether the program is a waste of resources, and of participant's time. Indeed, an evaluation might uncover the fact that the program has no beneficial effects, or even that the program is harmful! Once a program has been established as *beneficial*, it would likely be *unethical* to conduct an evaluation where the program is withheld from some participants. However, we could then consider a comparison of the program with an enhanced version of the program that might confer even more benefits.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

  programgroup["program group"] --> program

  program[program] --> outcome1[outcome]

  comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2 stroke:#000000,stroke-width:3px,font-size:36px,color:black;

```

We hope that our results will show that those participants who have participated in the program group have better outcomes than those who were members of the comparison group.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

  programgroup["program group"] --> program

  program[program] --> outcome1[better outcome]:::forestgreen

  comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2 stroke:#000000,stroke-width:3px,font-size:36px,color:black;

  classDef green fill:#80BD41,stroke:#000000,stroke-width:2px,color:#000000;
  
  classDef springgreen fill:#00FF7F,stroke:#000000,stroke-width:2px,color:#000000;
  
  classDef forestgreen fill:#CDE498,stroke:#000000,stroke-width:2px,color:#000000;
  
```

# Different Subgroups of Individuals 

One big question or concern in conducting an evaluation of this type is that our group of participants may be composed of different subgroups of individuals. These groups might be different in that they might repsresent individuals of different *racial*, *ethnic* or *gender* identities, might be people from different communities, or might have quite different sets of past experiences. 

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart TB

  subgraph study["study participants"]

  A["Group A: 100 people"]

  B["Group B: 100 people"]

  end

```

# Are Those In the program Group And Comparison Group Similar?

If participants from Group A and Group B are evenly distributed across the program and treatment groups, then we are not worried about the idea that an apparent effect of the program is because of unequal allocation of groups to the program.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

subgraph study["study participants"]

  A["Group A: 100 people"]

  B["Group B: 100 people"]

  end

  A --> |"roughly 50"| programgroup["program group"]

  B --> |"roughly 50"| programgroup["program group"]

  programgroup["program group"] --> program

  program[program] --> outcome1[outcome]

  A --> |"roughly 50"| comparisongroup

  B --> |"roughly 50"| comparisongroup

  comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2,3,4,5,6 stroke:#000000,stroke-width:3px,font-size:14px,color:red;

```

# Random Assignment

One way of accomplishing this *even distribution* would be by *randomly* assigning participants to the program.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

subgraph study["study participants"]

  A["Group A: 100 people"]

  B["Group B: 100 people"]

  end

  A --> |"roughly 50 RANDOMLY ASSIGNED"| programgroup["program group"]

  B --> |"roughly 50 RANDOMLY ASSIGNED"| programgroup["program group"]

  programgroup["program group"] --> program

  program[program] --> outcome1[outcome]

  A --> |"roughly 50 RANDOMLY ASSIGNED"| comparisongroup

  B --> |"roughly 50 RANDOMLY ASSIGNED"| comparisongroup

  comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2,3,4,5,6 stroke:#000000,stroke-width:3px,font-size:14px,color:red;

```

# Random Assignment Is Sometimes Not Possible

However, often the nature of the *program* is such that we want to allow *participants* in the study to *select their own level of participation*, or non-participation, in the program.

Often funders have objections to random assignment. Or the individuals or communities who are participating may have objections to random assignment.

# When Participants Choose Their Own Level Of Particpation

If *participants* select their own participation, then it is likely that many background characteristics of participants--such as their age, gender, socio-economic resources, and motivation--will affect their decision to participate, and their level of participation.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

participation{participation <br>decision}

age --> participation

gender --> participation

ses["socio-economic status"] --> participation

motivation --> participation

background --> participation

history --> participation

participation --> programgroup["program group"]

participation --> comparisongroup["comparison group"]

programgroup --> program

program[program] --> outcome1[outcome]

comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2,3,4,5,6,7,8,9,10 stroke:#000000,stroke-width:3px,font-size:36px,color:black;

```

If, for example, *more motivated individuals* and those with *more socio-economic resources* are the ones who choose to participate in the program, then any better outcomes might [**not**]{color="#FF0000"} be the result of the *program*, but a result of the *characteristics of individuals who chose to participate*.

```{mermaid}
---
config:
  look: handDrawn
  theme: default
---

flowchart LR

  programgroup["program group <br>(more motivated <br>better SES)"]:::forestgreen --> program

  program[program] --> outcome1[better outcome]:::forestgreen

  comparisongroup["comparison group"] ---> outcome0[outcome]

  linkStyle 0,1,2 stroke:#000000,stroke-width:3px,font-size:36px,color:black;

  classDef springgreen fill:#00FF7F,stroke:#000000,stroke-width:2px,color:#000000;
  
  classDef forestgreen fill:#CDE498,stroke:#000000,stroke-width:2px,color:#000000;
  
```

# What To Do?

*Accounting for*, *ruling out*, or *controlling for* these background factors is one of the major reasons we use statistics.

```{r}

program <- seq(1:6)

motivation <- c(2,1,0,7,6,5)

outcome <- motivation

mydata <- data.frame(program, motivation, outcome)

library(ggplot2)

ggplot(mydata,
       aes(x= program,
           y = outcome)) +
  geom_point() + 
  geom_smooth(method = "lm")

```




