---
title: "Patterns Of Caregiver Aggressive And Nonaggressive Discipline"
author: "TBD"
date: "today"
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
    highlight-style: haddock
---

```{r}
#| message: false
#| warning: false

library(readr) # read CSV

library(tidyr) # tidy data

library(dplyr) # data wrangling, and for pipe %>%

library(ggplot2) # beautiful graphs

library(ggforce) # extensions to ggplot

library(plotly) # interactive graphics

library(countrycode) # work with country names

```

# Background

Caregivers use a variety of disciplinary methods to respond to undesired child behavior. Many caregivers use nonaggressive forms of discipline, such as verbal reasoning and redirection. Some caregivers use aggressive forms of discipline, such as spanking and yelling. However, most caregivers use a combination of aggressive and nonaggressive discipline. To date, a disproportionately small number of caregiver discipline studies are conducted in low- and middle-income countries (LMICs), and few studies in low-resource contexts examine aggressive and nonaggressive behaviors simultaneously.

# Objective

This study aims to elucidate caregiver patterns of 11 disciplinary behaviors used in LMICs, and examine how these patterns relate to child outcomes and household characteristics.

# Participants and setting

Data came from the fourth and fifth rounds of UNICEF Multiple Indicator Cluster Surveys (MICS) distributed between 2009 and 2017 (N = 218,824 respondents across 63 countries). Focal children were 3–4 years old.

```{r}
#| echo: false
#| warning: false

country <- c("Algeria",  "Argentina",  "Bangladesh",  "Barbados",  "Belarus",  "Belize",  "Benin",  "Bosnia and Herzegovina",  "Cameroon",  "Central African Republic",  "Chad",  "Democratic Republic of the Congo",  "Republic of the Congo",  "Costa Rica",  "Cote d'Ivoire",  "Dominican Republic",  "El Salvador",  "Eswatini",  "Ghana",  "Guinea",  "Guinea Bissau",  "Guyana",  "Iraq",  "Jamaica",  "Kazakhstan",  "Kenya",  "Kosovo",  "Kyrgyzstan",  "Laos", "Macedonia",  "Madagascar",  "Malawi",  "Mali",  "Mauritania",  "Mexico",  "Moldova",  "Mongolia",  "Montenegro",  "Nepal",  "Nigeria",  "Pakistan",  "Panama",  "Paraguay",  "Sao Tome and Principe",  "Senegal",  "Serbia",  "Sierra Leone",  "Somalia",  "St. Lucia",  "State of Palestine",  "Suriname",  "Thailand",  "The Gambia",  "Togo",  "Trinidad and Tobago",  "Tunisia",  "Turkmenistan",  "Ukraine",  "Uruguay",  "Vietnam",  "Zimbabwe")

MICScountries <- data.frame(country)

MICScountries$country_iso <- countrycode(MICScountries$country, 
                                 'country.name', 
                                 'iso3c')

```

```{r}
#| echo: false

g <- list(showframe = TRUE, 
          showcoastlines = TRUE, 
          showcountries = TRUE,
          showland = TRUE,
          countrycolor = toRGB("grey"),
          showocean = TRUE,
          oceancolor = "white",
          projection = list(type = 'robinson' #,
                            # rotation = list(lon = -25,
                            #                 lat = 0,
                            #                 roll = 0)
                            ),
          showland = TRUE,
          landcolor = toRGB("lightgrey"),
          bgcolor = "transparent")

t <- list(family = "sans serif", 
          size = 8, 
          color = toRGB("black"))

l <- list(color = toRGB("black"), 
          width = 0.5)

p0 <- MICScountries %>% 
  plot_geo() %>% 
  add_trace(z = 1,
    text = ~country, 
    colors = "#1CABE2",
    hoverinfo = 'text',
    locations = ~country_iso, 
    showscale = FALSE,
    marker = list(size=10, line = l)) %>%
  layout(geo = g,
         plot_bgcolor='transparent',
         paper_bgcolor='transparent') %>% 
  config(displayModeBar = FALSE)

```

```{r}
#| echo: false
#| fig-cap: "Countries in UNICEF MICS Data"

p0

```

# Methods

Patterns of disciplinary behaviors were estimated using a multilevel latent class analysis (LCA). Multinomial regression analyses examined associations of disciplinary patterns with caregiver-reported child outcomes and household characteristics.

# Results

The LCA suggested caregiver discipline fell into three overall patterns: high behavioral control, moderate behavior control, and lower behavioral control. 

```{r}
#| message: false

patterns <- read_csv("patterns.csv")

patterns <- patterns %>%
  mutate(row_id = row_number())

```


```{r}

# library(scales)

# show_col(pal_viridis(option = "turbo")(15))

# data wrangling

patterns$Item <- factor(patterns$Item)

patterns$Item <- factor(patterns$Item, 
                        levels = c("Shouted at child", 
                                   "Called child names", 
                                   "Shook child", 
                                   "Spanked child",
                                   "Hit/slapped child on arm/leg",
                                   "Hit child with object",
                                   "Hit/slapped child on face/head/ears",
                                   "Beat as hard as one could",
                                   "Took privileges away from child",
                                   "Explained why behavior was wrong",
                                   "Gave child something else to do"))

# base graph

p0 <- patterns %>% 
  pivot_longer(!c(Item, row_id), 
               names_to = "discipline", 
               values_to = "value") %>%
  ggplot(aes(x = reorder(Item, row_id), 
             y = value,
             fill = Item)) + 
  # geom_point() +
  geom_col() +
  # facet_wrap(~discipline,
  #            ncol = 3) +
  coord_polar(theta = "x") +
  labs(title = "Patterns of Parental Discipline",
       x = "",
       y = "probability") +
  # scale_fill_viridis_d(option = "turbo") +
  scale_fill_manual(name = "Discipline", 
                    values = c("Shouted at child" = "#4147ADFF",
                               "Called child names" = "#4777EFFF", 
                               "Shook child" = "#39A4FBFF", 
                               "Spanked child" = "#FE9B2DFF",
                               "Hit/slapped child on arm/leg" = "#F46516FF",
                               "Hit child with object" = "#DB3A07FF",
                               "Hit/slapped child on face/head/ears" = "#B21A01FF",
                               "Beat as hard as one could" = "#7A0403FF",
                               "Took privileges away from child" = "#62FC6BFF",
                               "Explained why behavior was wrong" = "#A2FC3CFF",
                               "Gave child something else to do" = "#D2E935FF")) +
  theme_bw() +
  theme(axis.text.x = element_blank()) 

# p0 # replay

# faceted; all facets

p1 <- p0 + 
  facet_wrap(~discipline) + 
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 6))

p1 # replay

```


The lower behavioral control class was associated with the most advantageous child outcomes and household socio-demographic characteristics. 

```{r}

p1C <- p0 + 
  facet_wrap_paginate(~discipline,
                      ncol = 1,
                      nrow = 1,
                      page = 3) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 6))

p1C # replay

```

The high behavioral control class was associated with the most disadvantageous child outcomes and household characteristics.

```{r}
#| eval: false

p1B <- p0 + 
  facet_wrap_paginate(~discipline,
                      ncol = 1,
                      nrow = 1,
                      page = 2) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 6))

p1B # replay

```


```{r}

# facet by facet

p1A <- p0 + 
  facet_wrap_paginate(~discipline,
                      ncol = 1,
                      nrow = 1,
                      page = 1) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 6))

p1A # replay

```


# Conclusions

Efforts should be employed to reduce aggressive behaviors and promote positive parenting among caregivers in LMICs.

```{mermaid}

flowchart TB

  classDef yellow fill:#FFC20E,stroke:#000000,stroke-width:2px,color:#000000;
  
  classDef blue fill:#374EA2,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef green fill:#00833D,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef orange fill:#D86018,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  classDef red fill:#9A3324,stroke:#000000,stroke-width:2px,color:#FFFFFF;
  
  subgraph programs[interventions]
  
  interventions:::blue --> |"reduce"| aggressivebehavior[aggressive parenting]:::red
  
  interventions:::blue --> |"increase"| positiveparenting[positive parenting]:::green
  
  end
  
  programs --> |"improve"| outcomes[child outcomes]:::yellow

  linkStyle 0 stroke:#000000,stroke-width:3px,font-size:18px,color:red;
  
  linkStyle 1,2 stroke:#000000,stroke-width:3px,font-size:18px,color:green;
  
  style programs fill: #F2F2F2

```







